---
title: |
      **Inhomogeneity detection and correction on daily observational climate data series in Spain**
subtitle: 'Dataset, 1961-2020'
author: Santiago Beguería, Fergus Reig
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    df_print: paged 
    highlight: tango
    number_sections: yes
    theme: readable
    toc: yes
    fig_caption: yes
    toc_float: true
    collapsed: false
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 4
abstract: |
  .
---

# Introduction

```{r setup, include=FALSE}
# Global options
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Dependencies (and install packages, if needed)
if (!require('pacman')) install.packages('pacman')
pacman::p_load(sp, rgdal, chron, dplyr)

# Source snht functions - ESTO DEBERÍA SUSTITUIRSE POR INSTALAR EL PAQUETE
source('./functions_snht.r')
#library(devtools)
#install_github('lcsc/gap_filling@pkg_transform', auth_token = "XXXXXXXXXXXXXXX")

# Projections
prj_geo <- '+proj=longlat +datum=WGS84 +no_defs'
prj_utm <- '+proj=utm +zone=30 +ellps=intl +towgs84=-87,-98,-121,0,0,0,0 +units=m +no_defs'

# A shapefile with a map of Spain with provinces - FALTA UNO CON CANARIAS
provs_main <- readOGR('../data_raw/sig', layer='provs_main')
#provs_can <- readOGR('../data_raw/sig', layer='provs_can')
```


Standard Normal Homogeneity Test (SNHT)

The standard normal homogeneity test (SNHT) can be used to detect 
inhomogeneities in hydrological, climatological or any other time series
(Alexandersson 1986).

Three main functions are available (functions_snht.r):

* `snht()` detects a single change point in a test series and tests for
its statistical significance.
* `snht.Q()` can be used to build a test series suited to apply the SNHT
from one candidate time series and several auxiliary ones.
* `snht.daily()` uses the two functions above to apply the SNHT to a
daily time series. The function applies the procedure recursively and corrects
the found inhomogeneities, until no more significant change points are found.

There are two auxiliary functions:

* `snht.plot()` produces a diagnostic plot with the output of a call to
`snht()`.
* `snht.daily.plot()` produces diagnostic plots with the output of a call to
`snht.daily()`.


In this notebook we apply the procedure to time series of minimum temperature.


----

# Minimum temperature

The data consists on the series that were filled-in in the previous step of
the data process. Therefore, only a reduced number of series will be used.
Those series are all serially complete, or very close to 100%.

The test function must decide which series to use for constructing a
reference series for each candidate. The distance matrix represents a ranking
of the series, so the closest n stations will be used. Here we shall use the
correlation distance.


## Input data

```{r echo=FALSE}
# Read station metadata
sts <- read.table('../data_raw/data_coor/tmin.csv', sep=';', header=TRUE)
colnames(sts) <- c('id', 'name', 'alt', 'utm_x', 'utm_y', 'lon', 'lat', 'prov')
sts$prov <- as.factor(sts$prov)

# Read original (quality-controlled) data
dat_ori <- read.table(file='../data_raw/data_sort/tmin.csv.gz', header=TRUE,
                  sep=';', check.names=FALSE)
tmes_ori <- seq(from=as.Date(rownames(dat_ori)[1], format='%d/%b/%Y'),
            to=as.Date(rev(rownames(dat_ori))[1], format='%d/%b/%Y'),
            by='day')
dat_ori <- dat_ori %>% 
  dplyr::filter(tmes_ori >= as.Date('1961-01-01') &
                  tmes_ori <= as.Date('2022-12-31'))

# Read filled data
dat_fil <- read.table(file='../data/tmin_filled.csv.gz', header=TRUE,
                  sep=';', check.names=FALSE)

# Check the dimensions of the two matrices
nrow(dat_ori) ==  nrow(dat_fil) # must be TRUE

# Check that station identifiers correspond between the two objects
sum(!(colnames(dat_ori) %in% sts$id)) # must be 0
sum(!(colnames(dat_fil) %in% sts$id)) # must be 0

# Cut to common columns, and sort both objects equally
sts <- sts[sts$id %in% colnames(dat_fil),]
sts <- sts[order(sts$id),]
dat_fil <- dat_fil[, order(colnames(dat_fil))]
dat_ori <- dat_ori[, colnames(dat_fil)]
sum(names(dat_fil) != names(dat_ori)) # must be 0

# A vector of dates corresponding to the observations
tmes <- seq(as.Date('1961-01-01'), as.Date('2022-12-31'), by='day')
tmes_ym <- as.character(tmes, format='%Y-%m')
yeas <- as.character(tmes, format='%Y')
mons <- as.character(tmes, format='%m')
weeks <- as.character(tmes, format='%V')
days <- as.character(tmes, format='%d')

# Read or compute a distance matrix
f <- '../data/tmin_distance.RData'
if (!file.exists(f)) {
  
  # correlation distance matrix; reversed, so lower is better
  # correlation of candidate stations against all stations
  w <- sts$cand
  system.time(crr <- -cor(dat, dat[,w], use='pairwise'))
  #  sec
  crr[crr==-1] <- NA # nullify self correlations
  dim(crr)
  
  # make the matrix square
  dst <- matrix(NA, nrow=ncol(dat), ncol=ncol(dat))
  colnames(dst) <- rownames(dst) <- colnames(dat)
  dst[,w] <- crr
  
  save(dst, file=f)

} else {
  
  load(f)

}
# cut dst to the same stations in dat_fil
dst <- dst[, colnames(dat_fil)]

# Show stations in a map
stsSp <- sts
coordinates(stsSp) <- c('utm_x','utm_y')

plot(provs_main, axes=TRUE, cex.axis=0.75)
plot(stsSp, add=TRUE, col='red', pch=19, cex=0.25)
title('Min. temperature station network')
```




## Apply the test

The following code applies the SNHT test recursively to all the time series in
the data set.

```{r}
for (s in colnames(dat)) {
  
  # identify the series
  w <- which(colnames(dat) == s)
  id <- sts$id[w]
  nam <- paste0(id,', ', sts$name[w], ' (', sts$prov[w], ')' )
  
  # determine candidate series
  can <- dat %>% 
    dplyr::select(any_of(s))
  can <- can[,1]
  
  # determine auxiliary series (5 best correlated)
  aux <- dst %>% 
    as.data.frame() %>% 
    dplyr::select(any_of(s)) %>% 
    dplyr::filter(rownames(.) %in% colnames(dat)) %>% 
    dplyr::arrange(1) %>% 
    dplyr::filter(row_number() %in% 1:5)
  aux <- dat[,colnames(dat) %in% rownames(aux)]
  
  # apply SNHT test, and correction
  can_new <- snht.daily(can, aux, tmes, max.loops=10, method='diff',
                        excl=4*365, clevel=90, cl.factor=365,
                        corr='monthly', robust=FALSE)
  
  # plot the result
  pdf (paste0('../img/tmin/', id, '.pdf'))
    snht.daily.plot(can, can_new, tmes)
  dev.off()
}
```







